A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-20 tex/MercuryNumerical.tex
added new tex version

\usepackage[utf8]{inputenc}         % this is needed for Umlaute
\usepackage[english,ngerman]{babel} % this is needed for Umlaute
\usepackage[T1]{fontenc}            % this is needed for correct output of umlauts in pdf
\usepackage{cite}

\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage[%
  unicode=false,%
  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
  pdfauthor={C. Körber, I. Hammer, , J.-L. Wynen, C. Müller and C. Hanhart},%
  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.  The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
  colorlinks=true,        % true: colored links
  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
  citecolor=magenta,        % color of links to bibliography
  filecolor=YellowGreen,      % color of file links
  urlcolor=blue           % color of external links
]{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

@@

\usepackage{subcaption} % subfigures
\usepackage{caption}    % itemize in caption

@@

\author{
	C.~Körber$^{1}$,
	I.~Hammer$^{1}$,
	??$^{3}$,
	J.-L.~Wynen$^{1}$,
	C.~Müller$^{2}$ and 
	C.~Hanhart$^{1}$
}
\address{
	$^1$ \textit{Institute für Kernphysik and Institute for Advanced Simulations, Forschungszentrum Jülich, D-52425 Jülich, Germany}\\
	$^2$ \textit{??}\\
	$^2$ \textit{??}
}
\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

@@

\listoftodos

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that can be used to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \texttt{VPython}.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concepts of differential equations, their discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In particular tools are developed to extract the relevant quantity from the result of the simulation.
Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation are sensible.

We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 17 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories observed.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
Said differently: a force $\vec F$ shows up by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity.
We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newtons law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
This is what we need $r_S$ for later in the discussion.
That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
With this Newtons second law reads 
\begin{equation}
	\ddot{r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \, , \qquad
	\ddot{\vec r} =   \ddot{r} \, \frac{\vec{r}}{r} \, .
\label{eq:newton}
\end{equation}
In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
Depending on time available for the course it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class, it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable. \todo{@CH:citation here}

We allow for a movement of the perihelion by modifying the potential of Eq.~(\ref{eq:newton}) with the following ansatz
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \, \frac{\vec{r}}{r} \ ,
\label{eq:newton_art}
\end{equation}
where $\alpha$ is some parameter related to the effects by General Relativity.
The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
Replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r_{MS})\sim 10^{-7}$, suggests that it should be sufficient to include this kind of term to first order only.\footnote{Considering another order would lead qualitatively to the same result, since every change in the potential leads to a perihelion motion. Only the value of the pre-factor would change.}
For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
\begin{equation}
	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
\end{equation}
where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{@all: Acknowledgments}
Here they come

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{References}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{unsrt}%{alphadin}
\bibliography{References}

>>>>

\usepackage[utf8]{inputenc}         % this is needed for Umlaute
\usepackage[main=american]{babel} % this is needed for Umlaute
\usepackage[T1]{fontenc}            % this is needed for correct output of umlauts in pdf
\usepackage{cite}

\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage[%
  unicode=false,%
  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
  pdfauthor={C. K\"orber, I. Hammer, J.-L. Wynen, J. Heuer, C. M\"uller and C. Hanhart},%
  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.  The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
  colorlinks=true,        % true: colored links
  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
  citecolor=magenta,        % color of links to bibliography
  filecolor=YellowGreen,      % color of file links
  urlcolor=blue           % color of external links
]{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

@@

\usepackage{subcaption} % subfigures

@@

\author{
	C.~K\"orber$^{1}$,
	I.~Hammer$^{1}$,
	J.-L.~Wynen$^{1}$,
	J.~Heuer$^{2}$,
	C.~M\"uller$^{3}$ and 
	C.~Hanhart$^{1}$
}
\address{
	$^1$ \textit{Institute f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^3$ \textit{JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
}
\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

@@

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \texttt{VPython}.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their 
discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation are sensible.

We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülderakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24 
German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

A physical system is understood, if the assumed force leads to the trajectories observed.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
Said differently: a force $\vec F$ shows up by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity.
We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newtons law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
With this Newtons second law reads 
\begin{equation}
	\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, .
\label{eq:newton}
\end{equation}
In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded or to open orbits. 
In the latter case the planet simply disappears from the sun, since a given force can capture only bodies with not too large momenta.
The students may study those scenarios within their simulations  by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter. 
However, besides $r_S$, which is the characteristic length scale for the gravitational potential
of the sun, one can define one more parameter, which is characteristic for the dynamics of mercury, namely
\begin{equation}
h=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $h$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
for the modified potential
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{h}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand $h$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
We thus find that $(h/ r_{MS}^2) = 4\times10^{-6}$. Furthermore $(r_S/r_{MS})=6\times 10^{-6}$. 
Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $h$ or $r_S$
should be suppressed by additional six orders of magnitude and thus can be safely neglected.
The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
theory --- here General relativity. However, since we have only a single number to study, the angular motion of the perihelion,
we can only fix either $\alpha$ or $\beta$.

%For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
%Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
%\begin{equation}
%	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
%\end{equation}
%where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{acknowledgments}

This work is supported in part by NSFC and DFG through funds provided to the
Sino--German CRC110 ``Symmetries and the Emergence of Structure in QCD''.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{References}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{References}
\bibliographystyle{unsrt}%{alphadin}
\bibliography{References}
